Evaluation of Pre-Trained Models for Many-to-English Translation

### 2024-08-04 paper/paper.tex
updated

% Packages
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[backend=bibtex]{biblatex}
\usepackage{comment}
\usepackage{hyperref}

@@

\title{Evaluation of Pre-Trained Models for Translation in Different Languages}
\author{Stefan Liemawan Adji}
\date{\today}

@@

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, NLLB \cite{nllb200-2020}, M2M \cite{fan-2020-m2m100}, and PolyLM \cite{wei-2023-polylm}. Most of these models are multilingually trained, able to translate between any of the supported pair of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Pre-trained models are often not evaluated under the same datasets or benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}. Therefore, the performance of these models across different languages seem to be yet unclear.

Through simple experimentation, this paper aims to evaluate the performance of existing pre-trained models (PTMs) across various languages. Without any training, the study involves running inference for translations from different languages into English. A dataset curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 sentence pairs across 14 source languages, is utilised. The evaluation includes a one-to-one PTM, OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs, such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is compared using the BLEU score \cite{papieni-2002-bleu}.

@@

Since LLMs and pre-trained models require vast amounts of training data, the availability of parallel corpora is crucial for building effective models. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

@@

\begin{comment}
PolyLM \cite{wei-2023-polylm} (32 languages), and
mT5 \cite{xue-2021-mt5} (101 languages).

\subsection{Sequence-to-Sequence Models}

GPT \cite{openai-2024-gpt4}

mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages. However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.

T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.

MistralAI \cite{jiang-2023-mistral-7b}

\end{comment}

@@

READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)

@@

>>>>

% Packages
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[backend=bibtex]{biblatex}
\usepackage{comment}
\usepackage{hyperref}
\usepackage{amsmath}

@@

\title{Evaluation of Pre-Trained Models for Many-to-English Translation}
\author{Stefan Liemawan Adji}
\date{\today}

@@

Since the introduction of Transformers in 2017 \cite{vaswani-2017-attention}, Natural Language Processing (NLP) and machine translation in particular reached a giant milestone. The following years saw the birth of Large Language Models (LLMs) such as BERT \cite{devlin-2019-bert}, GPT \cite{openai-2024-gpt4}, and T5 \cite{raffel-2023-t5}, which revolutionised both MT and the whole field of NLP. Then in early 2020s, several pre-trained models (PTM) that are specifically designed for machine translation emerged, namely mBART \cite{liu-2020-mbart}, mT5 \cite{xue-2021-mt5}, NLLB \cite{nllb200-2020}, M2M \cite{fan-2020-m2m100}, and PolyLM \cite{wei-2023-polylm}. Most of these models are multilingually trained, allowing for many-to-many translation: able to translate between any of the supported pair of languages. This allows the models to generalise over shared lexical and linguistic among languages, and have been shown to increase performance compared to one-to-one translation models \cite{liu-2020-mbart}.

Intento published 'The State of Machine Translation 2024' \cite{intento-2020} providing an in-depth evaluation of popular MT engines and LLMs. However, the biggest drawback in this report is that the selected LLMs are general LLMs such as GPT, LLaMa, Mistral, instead of MT-specific LLMs.

Despite these advancements, pre-trained models are often evaluated using different set of benchmarks \cite{liu-2020-mbart,nllb200-2020,fan-2020-m2m100,wei-2023-polylm}, making it difficult to gauge their relative effectiveness across various languages. Nevertheless, there does not seem to be much work on comparing or benchmarking different pre-trained models in machine translation.

Through simple experimentations, this paper aims to evaluate the performance of existing pre-trained models (PTMs) on many-to-English translation across 14 source languages. Although fine-tuning multilingual PTMs has been proven to increase model performance \cite{cooper-stickland-2021-recipes}, no pre-training or fine-tuning is performed in this study for simplicity reasons. A dataset is curated from the Tatoeba repository \cite{tatoeba}, containing 1,241 parallel sentence pairs across source and target languages. The models includes a one-to-one PTMs: OPUS-MT \cite{tiedemann-2020-opus-mt}, and multilingual PTMs: such as mBART-50 \cite{liu-2020-mbart}, NLLB-200 \cite{nllb200-2020}, and M2M-100 \cite{fan-2020-m2m100}. The performance of these models is evaluated using the BLEU score \cite{papieni-2002-bleu}.

% (Multilinguality has been explored in the supervised NMT literature, where it has been shown to enable information sharing among related languages.) \cite{garcia-2020-multilingual}
% (NMT systems have lower translation quality on very long sentences) \cite{koehn-2017-challenges}

@@

Since Neural Machine Translation (NMT) systems require vast amounts of training data, the availability of parallel corpora is crucial for building effective models \cite{koehn-2017-challenges}. The lack of extensive parallel corpora, especially for low-resource languages, leads to suboptimal performance in NMT techniques compared to their high-resource counterparts \cite{ranathunga-2023-nmt-low-res}. OPUS \cite{opus} is a comprehensive collection of open-source parallel corpora used extensively in the field of machine translation (MT). It includes corpora for 744 languages and contains over 1,210 different datasets, amassing a total of 45,945,946,108 sentence pairs. Tatoeba \cite{tatoeba} is another prominent resource in the field of MT and NLP, known for its extensive collection of translated sentences. As of July 2024, it contains 12,186,207 sentences over 423 supported languages, growing daily through volunteer contributions.

The Tatoeba Challenge \cite{tiedemann-2020-tatoeba-challenge}, IWSLT \cite{agarwal-iwstl-2023}, TED \cite{ye-2018-word-embeddings-ted}, Flores \cite{goyal-2022-flores}, and WMT \cite{barrault-2020-wmt} are among the biggest and most popular parallel corpora datasets, allowing for many-to-many translations.

Despite being used diligently by Meta AI papers for benchmarking \cite{fan-2020-m2m100,nllb200-2020}, these datasets are not seen used

@@

mRASP2 \cite{pan-2021-mrasp2}

\subsection{Evaluation Metrics}

Bilingual Evaluation Understudy (BLEU) \cite{papieni-2002-bleu} is the most commonly used metrics for machine translation (MT). It assesses how well a candidate translation matches the reference translation using precision metrics for n-grams and incorporates a brevity penalty to prevent overly short translations from achieving high scores.

The n-gram precision, as presented in the original BLEU paper \cite{papieni-2002-bleu}, is calculated as:

\begin{equation}
    p_n = \frac{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}_{\text{clip}}(\text{n-gram})}{\sum_{C \in \{Candidates\}} \sum_{\text{n-gram} \in C} \text{Count}(\text{n-gram})}
\end{equation}

Where:
\begin{itemize}
    \item \( p_n \) is the precision for n-grams.
    \item \( \sum_{C \in \{Candidates\}} \) denotes the summation over all candidate translations.
    \item \( \sum_{\text{n-gram} \in C} \) denotes the summation over all n-grams in a candidate translation \( C \).
    \item \( \text{Count}_{\text{clip}}(\text{n-gram}) \) is the clipped count of the n-gram, which is the count of the n-gram in the candidate translation limited by the maximum count of that n-gram in any reference translation.
    \item \( \text{Count}(\text{n-gram}) \) is the count of the n-gram in the candidate translation.
\end{itemize}

Thus, the BLEU score is calculated as:

\begin{equation}
    \text{BLEU} = BP \cdot \exp \left( \sum_{n=1}^{N} w_n \log p_n \right)
\end{equation}

Where:
\begin{itemize}
    \item \( BP \) is the brevity penalty.
    \item \( p_n \) is the precision for n-grams.
    \item \( w_n \) is the weight for each n-gram (often uniformly distributed, so \( w_n = \frac{1}{N} \)).
\end{itemize}

The brevity penalty (BP) is calculated as:

\begin{equation}
    BP = \begin{cases}
        1                     & \text{if } c > r    \\
        e^{(1 - \frac{r}{c})} & \text{if } c \leq r
    \end{cases}
\end{equation}

Where:
\begin{itemize}
    \item \( c \) is the length of the candidate translation.
    \item \( r \) is the effective reference length.
\end{itemize}

While other metrics exist for machine translation such as METEOR \cite{lavie-2007-meteor} and COMET \cite{rei-2020-comet}, it is far more practical to implement BLEU due to its popular usage in other works.

% Introduce the concept of evaluation metrics for MT, explaining why BLEU is chosen for this study and if there are other metrics that could complement it (like METEOR or ROUGE).

% \begin{comment}
% PolyLM \cite{wei-2023-polylm} (32 languages), and
% mT5 \cite{xue-2021-mt5} (101 languages).

% \subsection{Sequence-to-Sequence Models}

% GPT \cite{openai-2024-gpt4}

% mBERT \cite{devlin-2019-bert} is the multilingual version of BERT, introduced in the same paper, trained on 104 different languages. However, the model been shown to suffer on low-resource languages \cite{wu-2020-mbert-are-all}.

% T5 \cite{raffel-2023-t5} is a Text-to-Text Transfer Transformer model trained on English.

% MistralAI \cite{jiang-2023-mistral-7b}

% \end{comment}

@@

\begin{figure}[htbp]
    \centering
    \includegraphics[width=1\linewidth]{figures/word_count_box.png}
    \caption{Dataset word count distribution, consisting of mostly short sentences.}
    \label{fig:word_count_box}
\end{figure}

Figure \ref{fig:word_count_box} shows a box plot of sentences word count. It can be seen that the majority of sentences are short.

Chinese and Japanese are counted per letter.

While most languages cluster around 4-6 words per sentence, there are notable exceptions like Japanese, which exhibits much less variability. This analysis can be useful for understanding language-specific characteristics in sentence structure, which could inform tasks like translation, text processing, or linguistic studies.

@@

% READ THIS \cite{intento-2020} (exactly what this paper should do, maybe compare results and find insights)
